### 2016-07-07 Projects/DelCo_2012/Report/DelCo_2012.tex
Text mod through 3.2
Minor adds and typos fixed.  Re-ordered fuel loads

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and firefighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service complements the NIST research program on firefighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural firefighting with the objective of enhancing the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural firefighting.

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAF for structural firefighting.

This paper documents Task 3 of the project. NIST conducted several series of experiments to provide scientific measurements and comparisons between CAFS and water-only hose streams for different compartment fire scenarios.

Three types of experiments were conducted:
\begin{enumerate}
\item spray density measures under non-fire conditions
\item gas layer cooling
\item fire suppression in experimental structures
\end{enumerate}

The key measures were to understand the cooling and suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAF during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Departments across the country began to experiment with CAFS systems for use in structural firefighting. In most documented cases, the findings about the effectiveness of the CAFS were favorable, yet after decades of fire department trials, CAF has still not been significantly incorporated into structural firefighting tactics. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAFS inside a structure.

@@

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. That study examined a broad range of tests in order to determine those parameters that most critically effect fire-fighting performance.  The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}:

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating wood based fuels, mass retention on the wood fuels and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhance the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allows the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. The catalyst for this study was as a result of this order.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package was composed mainly of wood based materials, as a way to increase the challenge of the suppression experiment. After the wood based fuel package experiments were completed, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression. Baseline operating conditions were agreed upon at the workshop for these experiments (Table~\ref{tab:op_condition}). Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.

@@

Thirty-two tests were conducted to determine spray density measures based on various nozzle patterns and locations. The gas cooling and fire suppression tests simulated an interior fire attack through the deployment of hose lines inside the structures with fixed nozzles. Therefore, the purpose of these tests was to determine the appropriate nozzle locations for the gas cooling and fire suppression tests to follow by gaining an understanding of the water distribution throughout the structures.

@@

Eighty-eight experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on interior conditions throughout the structure when suppression tactics were applied to a room adjacent to the fire room. Detailed discussion of the tests and results are in Section~\ref{sec:Gas_Cooling}.

@@

The burn room where the fuel load was located measured 3.8~m (12.5~ft) by 5.7~m (18.7~ft) with a ceiling height of 3.35~m (11.0~ft). The adjacent room where the water was applied for gas cooling measured 4.1~m (13.4~ft) by 5.7~m (18.7~ft), also with a ceiling height of 3.35~m (11.0~ft). The open door from the adjacent room to the exterior measured 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide.

@@

Note, that as the pallets burned away and the hot gas layer temperatures decreased, the steel shutters on the window to the burn room were opened and additional pallets were added. Pallets were added to the existing fuel locations until the flames from the piles reached the ceiling of the burn room again. The number of pallets added each time varied.

@@

@@

The fuel load for these experiments consisted of one bale of hay, twelve pallets, eleven and a half sheets plywood on the walls and the wood ``floor assembly". Two stacks of pallets with hay were used as the first items ignited in each experiment. Each stack was composed of six pallets with a half bale of hay. The hay was layered between each pallet. Figure~\ref{fig:Wood_Fuel_Load} shows the pallets prior to ignition. The pallets and hay were weighed prior to each experiment. For the four pallet experiments, the average mass of the two stacks of pallets and the bale of hay was 232.4~kg (511.3~lb).
\begin{figure}[!ht]
	\includegraphics[width=0.65\columnwidth]{../Figures/Pictures/Wood_Fuel_Package}
	\caption{Photograph of Southeast corner of burn room with wood fuel load.}
	\label{fig:Wood_Fuel_Load}
\end{figure}

@@

The fuel load for these experiments consisted of 2 sleeper sofas and 2 chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Fig.~\ref{fig:Furniture_Fuel_Load}. The couches and chairs are setup to mimic a typical seating area with one couch against the south and east wall respectively with each being 0.9~m (36~in) away from the corner. A chair is placed adjacent to each couch and the specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336 kg of fuel consisting of wood, polyurethane (PU), polyester (PET),and polypropylene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

\begin{sidewaystable}[!ht]
	\centering
	\scriptsize
	\caption{Fire Suppression Fuel masses}
	\renewcommand{\tabcolsep}{1pt}
	\begin{tabular}{lllllcc}
		\toprule[1.5pt]
		Item               & Component		& Quantity		&  Material Description             			&  Dimensions (m)            	&  Mass (kg)  		& Total Mass (kg) \\
		\midrule
		2-seat Sofa        &				& 2				& 85~\% PU Foam, 15~\% PET Fiber, Wood Frame 	&  1.68 L X 0.90 D X 0.86 H  	&  72.4    			& 144.8 \\
		           	 	   & Mattress	    & 1 per sofa	& Cover: 52~\% PP, 48~\% PET,                   &  1.83 L X 1.21 W X 0.14 H     &  14.6             & \\
		           	 	   &                &               & Filling: 66~\% PET Pad, 34~\% PET Batting     &                               &                   & \\
			           	   & Cushion    	& 2 per sofa	& 85~\% PU Foam, 15~\% PET Fiber   				&  0.69 L X 0.58 D X 0.14 H  	&  2.2     			& \\
		Blue Chair         &				& 2				& 85~\% PU Foam, 15~\% PET Fiber, Wood Frame   	&  0.81 L X 0.75 D X 0.91 H  	&  17.4    			& 34.8 \\
		    			   & Cushion        & 1 per chair   & 85~\% PU Foam, 15~\% PET Fiber				&  0.51 L X 0.57 D X 0.14 H  	&  1.7    			& \\
		Blue/Green Chair   & 				& 2				& 85~\% PU Foam, 15~\% PET Fiber, Wood Frame    &  0.77 L X 0.71 D X 0.89 H  	&  16    			& 32 \\
		                   & Cushion    	& 1 per chair	& 85~\% PU Foam, 15~\% PET Fiber		        &  0.55 L X 0.53 D X 0.15 H  	&  1.6    			& \\
		Paneling           &				& 8 panels		& Medium density fiberboard                     &  2.44 L X 1.22 W X 3.2~mm H   &  9    			& 72 \\
		Carpeting          &				& 22.3 m$^2$	& 100~\% PET                  					&  3.68 L X 6.05 W  			&  32.2				& 32.2 \\
		Carpet Padding     &				& 22.3 m$^2$	& PU                       	 				    &  3.68 L X 6.05 W X 11~mm H    &  19.8			    & 19.8 \\
		                   &                &               &                                               &                               & Total Mass        & 335.6 \\
		\bottomrule[1.25pt]
	\end{tabular}
	\label{tab:Fire_Suppression_Fuel_Masses}
\end{sidewaystable}

@@

@@

The fuel load for two of the experiments consisted of three sofas with an average mass of 48.7~kg; 51.4~m$^2$ (559~ft$^2$) of carpet and padding; and approximately 38 sheets of oriented strand board (OSB) paneling with dimensions 16~mm (0.63~in) x 1.2~m (4~ft) x 2.4~m (8~ft) along the East, West, and South walls and ceiling. The three couches are aligned along the West wall and spaced 1.2~m (4~ft) apart. The specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:furniture_2story}. In total, this fuel package was composed of wood, polyurethane (PU), polyester (PET),and polyprolene (PP).

@@

@@

Spray density experiments were conducted by placing 0.23~m$^2$ (2.5~ft$^2$) interlocking collector bins across the floor surface area of the burn rooms within each structure (cf. Fig~\ref{fig:water_buckets}). The buildings were sheathed in plastic for the spray density tests to protect and preserve as much of the structure as possible before any fire suppression tests were conducted. Water flowed for a specified duration and upon completion the bins were removed from the structure. Each bin was then weighed to quantify the impact of nozzle location and spray pattern. Thirty-three spray density tests were conducted; 21 were in the burn building for the gas cooling tests and 12 were in the concrete structure for the suppression experiments. The flow rate for each test was 120~gpm for water and 120~gpm plus 60~cfm for CAF. Tables\ref{tab:spray_density_tests} and \ref{tab:spray_density_tests2} provide an overview of each experiment. In the table, GC stands for the gas cooling experiments while SE stands for the suppression experiments. For all of the experiments, 100~ft (30~m) of 1~3/4~inch hoseline connected to a Blitz Fire monitor with a Metro 1 nozzle was used.

@@

For the spray density tests, Table~\ref{tab:spray_density_tests}, the mid position had center line of the stream aimed at 2.44~m (8~ft) south of the North wall and 2.13~m (7~ft) west of the East wall in the concrete burn building.  The back position had the center line of the stream aimed at 3.36~m (11~ft) south of the North wall and 2.13~m west (7~ft) of the East wall in the concrete burn building. Figure~\ref{fig:stream_position} shows the streams from the monitor nozzle for the middle and back positions respectively.

@@

After weighing each of the bins the mass of water (kg) in each bin was plotted by position to visualize the spray distribution. Figure~\ref{fig:Burn_Building_Test_1} shows that with a straight stream in the back position, the majority of the water accumulated along the North and East walls of the gas cooling room within the burn building (cf. Fig.~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}). Figure~\ref{fig:Burn_Building_Test_7} shows difference a fog stream as the distribution of water is far more dispersed around the gas cooling room. The remainder of the burn building spray density figures are included in Appnedix~\ref{app:spray_density}.

\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Bars/BB1}
	\caption{Gas Cooling Spray Density Test 1}
	\label{fig:Burn_Building_Test_1}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Bars/BB7}
	\caption{Gas Cooling Spray Density Test 7}
	\label{fig:Burn_Building_Test_7}
\end{figure}

>>>>

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and firefighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service complements the NIST research program on firefighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural firefighting with the objective of enhancing the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural firefighting.  This project was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

This paper documents Task 3 of the project. NIST conducted several series of experiments to provide scientific measurements and comparisons between CAFS and water-only hose streams for four different compartment fire scenarios.  Three different types of experiments were conducted:
\begin{enumerate}
\item spray density measures under non-fire conditions
\item gas layer cooling
\item fire suppression in experimental structures
\end{enumerate}

The key measures were to understand the gas cooling and fire suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAF during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel, and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS systems for use in structural firefighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable, yet after decades of fire department trials, CAF has still not been widely incorporated into structural firefighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

@@

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. That study examined a broad range of tests in order to determine those parameters that most critically effect fire-fighting performance.  The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating wood based fuels, mass retention on the wood fuels and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was the catalyst for this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood based materials.  The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished.  Wood based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.  The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure.  When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure.  This was done to minimize, the build-up of "dry" foam or a slug upon the initial flow of the nozzle into the test structure.

@@

Thirty-two tests were conducted to determine spray density measures based on various nozzle patterns and locations. The gas cooling and fire suppression tests simulated an interior fire attack through the deployment of hose lines inside the structures with fixed nozzles. The purpose of these preliminary tests was to determine the appropriate nozzle locations for the gas cooling and fire suppression tests to follow by gaining an understanding of the water distribution throughout the structures.

@@

Eighty-eight experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room.  Suppression agents were not applied to the fire in the burn room.  Detailed discussion of the tests and results are in Section~\ref{sec:Gas_Cooling}.

@@

The burn room where the fuel load was located measured 3.8~m (12.5~ft) by 5.7~m (18.7~ft) with a ceiling height of 3.35~m (11.0~ft). The adjacent room where the water was applied for gas cooling measured 4.1~m (13.4~ft) by 5.7~m (18.7~ft), also with a ceiling height of 3.35~m (11.0~ft). The open doorway that connected the burn room to the adjacent room and the doorway from the adjacent room to the exterior both measured 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide.

@@

Note, that as the pallets burned away and the hot gas layer temperatures decreased, the steel shutters on the window to the burn room were opened and additional pallets were added. Pallets were added to the existing fuel locations until the flames from the piles reached the ceiling of the burn room again. The shutters on the window would be then closed and the temperature of the hot gas layer in the adjacent room would be monitored in preparation for starting another gas cooling experiment.  The number of pallets added each time varied.

@@

The fuel load for experiments 1 through 8 consisted of 2 sleeper sofas and 2 chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Fig.~\ref{fig:Furniture_Fuel_Load}. The couches and chairs are setup to mimic a typical seating area with one couch against the south and east wall respectively with each being 0.9~m (36~in) away from the corner. A chair is placed adjacent to each couch and the specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336 kg (739 lbs) of fuel consisting of wood, flexible polyurethane (PU), polyester (PE),and polypropylene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

\begin{sidewaystable}[!ht]
	\centering
	\scriptsize
	\caption{Fire Suppression Furniture Fuel Package Description}
	\renewcommand{\tabcolsep}{1pt}
	\begin{tabular}{lllllcc}
		\toprule[1.5pt]
		Item               & Component		& Quantity		&  Material Description             			&  Dimensions (m)            	&  Mass (kg)  		& Total Mass (kg) \\
		\midrule
		2-seat Sofa        &				& 2				& 85~\% PU Foam, 15~\% PE Fiber, Wood Frame 	&  1.68 L X 0.90 D X 0.86 H  	&  72.4    			& 144.8 \\
		           	 	   & Mattress	    & 1 per sofa	& Cover: 52~\% PP, 48~\% PE,                    &  1.83 L X 1.21 W X 0.14 H     &  14.6             & \\
		           	 	   &                &               & Filling: 66~\% PE Pad, 34~\% PE Batting       &                               &                   & \\
			           	   & Cushion    	& 2 per sofa	& 85~\% PU Foam, 15~\% PE Fiber   				&  0.69 L X 0.58 D X 0.14 H  	&  2.2     			& \\
		Blue Chair         &				& 2				& 85~\% PU Foam, 15~\% PE Fiber, Wood Frame   	&  0.81 L X 0.75 D X 0.91 H  	&  17.4    			& 34.8 \\
		    			   & Cushion        & 1 per chair   & 85~\% PU Foam, 15~\% PE Fiber	    			&  0.51 L X 0.57 D X 0.14 H  	&  1.7    			& \\
		Blue/Green Chair   & 				& 2				& 85~\% PU Foam, 15~\% PE Fiber, Wood Frame     &  0.77 L X 0.71 D X 0.89 H  	&  16    			& 32 \\
		                   & Cushion    	& 1 per chair	& 85~\% PU Foam, 15~\% PE Fiber 		        &  0.55 L X 0.53 D X 0.15 H  	&  1.6    			& \\
		Paneling           &				& 8 panels		& Medium density fiberboard                     &  2.44 L X 1.22 W X 3.2~mm H   &  9    			& 72 \\
		Carpeting          &				& 22.3 m$^2$	& 100~\% Olefin fiber w/PP backing		     	&  3.68 L X 6.05 W  			&  32.2				& 32.2 \\
		Carpet Padding     &				& 22.3 m$^2$	& PU Foam                  	 				    &  3.68 L X 6.05 W X 11~mm H    &  19.8			    & 19.8 \\
		                   &                &               &                                               &                               & Total Mass        & 335.6 \\
		\bottomrule[1.25pt]
	\end{tabular}
	\label{tab:Fire_Suppression_Fuel_Masses}
\end{sidewaystable}

@@

The fuel load for experiments 11 through 14 consisted of one bale of hay, twelve pallets, eleven and a half sheets plywood on the walls and the wood ``floor assembly". Two stacks of pallets with hay were used as the first items ignited in each experiment. Each stack was composed of six pallets with a half bale of hay. The hay was layered between each pallet. Figure~\ref{fig:Wood_Fuel_Load} shows the pallets prior to ignition. The pallets and hay were weighed prior to each experiment. For the four pallet experiments, the average mass of the two stacks of pallets and the bale of hay was 232.4~kg (511.3~lb).
\begin{figure}[!ht]
	\includegraphics[width=0.65\columnwidth]{../Figures/Pictures/Wood_Fuel_Package}
	\caption{Photograph of Southeast corner of burn room with wood fuel load.}
	\label{fig:Wood_Fuel_Load}
\end{figure}

@@

\subsection{Attic Fires}

The fuel load for the attic fires, experiments 9 and 10, was essentially the OSB roof underlayment, and the wood trusses that support the roof.  A 1.2 m (4 ft) square hole was opened between the center of the southern room in the single story structure and the attic space. A single stack of pallets as described for the wood fuel package tests was loacted below the hole and ignited.  The fire was allowed to spread into the attic space.  In addition to the connecting hole between the room and the attic space, the attic had openings along the eave line on both sides of the structure and gable openings on each end of the attic space.

@@

In June of 2015, the larger compartment experiments, experiments 15 through 22, were conducted in the lower level of the two story structure.  Two different fuel packages were utitlized, the difference between the furniture fuel package used for experiments 15 and 16, and the furnitiure and pallet fuel package used in experiments 17 through 22 was the addition of two stacks of pallets.

@@

The fuel load for two of the experiments consisted of three sofas with an average mass of 48.7~kg; 51.4~m$^2$ (559~ft$^2$) of carpet and padding; and approximately 38 sheets of oriented strand board (OSB) paneling with dimensions 16~mm (0.63~in) x 1.2~m (4~ft) x 2.4~m (8~ft) along the East, West, and South walls and ceiling. The three couches are aligned along the West wall and spaced 1.2~m (4~ft) apart. The specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:furniture_2story}. In total, this fuel package was composed of wood, flexible polyurethane (PU), polyester (PE),and polyprolene (PP).

@@

The results are arranged in the order that the experiments were conducted.  For the fire suppression experiments, the fuel load and resulting fire hazard were intended to increase as the different structures and fuel packages were utilitized.

@@

Spray density experiments were conducted by placing 0.23~m$^2$ (2.5~ft$^2$) interlocking water colection pans across the floor surface area of the burn rooms within each structure (cf. Fig~\ref{fig:water_buckets}). The depth of the pans was 0.3 m (1 ft).  Water or CAF was flowed for a specified duration or until it appeared that a pan might be filled.  Upon completion the bins were removed from the structure. Each bin was then weighed to quantify the impact of nozzle location and spray pattern. Thirty-three spray density tests were conducted; 21 were in the burn building for the gas cooling tests and 12 were in the single story structure for the suppression experiments. The interior of the fire room in the single story structure was sheathed in plastic for the spray density tests to protect the gypsum board from being soaked with water prior to the fire suppression experiments.  The flow rate for each test was 120~gpm for water and 120~gpm plus 60~cfm for CAF. Tables\ref{tab:spray_density_tests} and \ref{tab:spray_density_tests2} provide an overview of each experiment. In the table, GC stands for the gas cooling experiments while SE stands for the suppression experiments. For all of the experiments, 100~ft (30~m) of 1~3/4~inch hoseline connected to a Blitz Fire monitor with a Metro 1 nozzle was used.

@@

For the spray density tests, Table~\ref{tab:spray_density_tests}, the mid position had center line of the stream aimed at 2.44~m (8~ft) south of the North wall and 2.13~m (7~ft) west of the East wall in the concrete burn building.  The back position had the center line of the stream aimed at 3.36~m (11~ft) south of the North wall and 2.13~m west (7~ft) of the East wall in the concrete burn building. Figure~\ref{fig:stream_position} shows the streams from the monitor nozzle for the middle and back positions respectively.  From the figures the qualitative difference of the nozzle positions can be seen.  When the stream is in the middle position, after the stream hits the ceiling, the impact region is relatively narrow with the majority of the water hitting the walls of the corner opposite the nozzle position.  When the nozzle is in the back position, the stream hits the ceiling at a much steeper angle and the droplets from the broken stream have a broader distribution throughout the room.

@@

After weighing each of the bins the mass of water (kg) in each bin was plotted by position to visualize the spray distribution. Figure~\ref{fig:Burn_Building_Test_1} shows that with a straight stream in the back position, the majority of the water accumulated along the North and East walls of the gas cooling room within the burn building (cf. Fig.~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}). Figure~\ref{fig:Burn_Building_Test_7} shows difference a fog stream as the distribution of water is far more dispersed around the gas cooling room. The remainder of the burn building spray density figures are included in Appendix~\ref{app:spray_density}.

\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Bars/BB1}
	\caption{Gas Cooling Spray Density Test 1, Straight Stream/Back nozzle position}
	\label{fig:Burn_Building_Test_1}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Bars/BB7}
	\caption{Gas Cooling Spray Density Test 7, 30 $^{\circ}$ Fog 
	\label{fig:Burn_Building_Test_7}
\end{figure}
